The Poisson approximate likelihood compared to the particle filter

### 2024-07-25 ms.Rnw
add plot

\documentclass[10pt]{article}
\usepackage{color}
\usepackage{graphicx} % Required for inserting 
%%\usepackage{algpseudocode} % Include the package for algorithmic environment
\usepackage{algorithm}
\usepackage{booktabs} % For better-quality horizontal lines
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{threeparttable} % For table notes
\usepackage{array} % For table formatting

@@

\usepackage{multirow}
\usepackage{fourier} 
\usepackage{booktabs}
\usepackage{array}
\usepackage{makecell}
\usepackage[font={small,it}]{caption}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{subcaption}
\usepackage{tabstackengine}
\usepackage[toc,page]{appendix}
\usepackage{graphicx, fullpage, verbatim, amsmath}
\usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
\usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
\usepackage{dsfont}
\usepackage{amsfonts}
\usepackage{bm}

@@

% Useful packages
\usepackage{graphicx}
\bibliographystyle{apalike}
\usepackage[colorlinks=true,citecolor=blue]{hyperref}
\newcommand\choleraDeath{\delta_{C}}

@@

@@

bake(file = paste0(out_dir, "round_01/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
  mifs_global_01
})

bake(file = paste0(out_dir, "round_01/ovov_el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

el_01 <- readRDS(paste0(out_dir,"round_01/ovov_el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"round_02/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0(out_dir, "round_02/ovov_el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

el_02 <- readRDS(paste0(out_dir, "round_02/ovov_el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "round_03/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0(out_dir,"round_03/ovov_el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
mif_el <- readRDS(paste0(out_dir,"round_03/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

bake(file=paste0(out_dir, "pfilter/ovov_sim_100.rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/pfilter_ovov_sim_100.rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/ovov_sim_100.rds"))
prop <- 420

@@

bake(file=paste0(out_dir, "pfilter/palfilter_ovov_100.rds"),{
  y <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    realdat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    y[i] <- lik_list$log_lik
  }
  y
})

@@

@@

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{|c c c|} 
 \hline
  & PAL & SMC  \\
  \hline
 OvOv & -6892.168 & -7200.866 \\ 
 \hline
\end{tabular}
\end{table}

>>>>

\documentclass[10pt]{article}
\usepackage{color}
\usepackage{graphicx} % Required for inserting 
%%\usepackage{algpseudocode} % Include the package for algorithmic environment
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{threeparttable} % For table notes
\usepackage{array} % For table formatting

@@

\usepackage{multirow}
\usepackage{fourier} 
\usepackage{booktabs}
\usepackage{array}
\usepackage{makecell}
\usepackage[font={small,it}]{caption}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{subcaption}
\usepackage{tabstackengine}
\usepackage[toc,page]{appendix}
\usepackage{graphicx, fullpage, verbatim, amsmath}
\usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
\usepackage{placeins, listings, textcomp, mathtools, multicol, tikz}
\usepackage{bm}

@@

% Useful packages
\usepackage{graphicx}
\bibliographystyle{apalike}
\usepackage[colorlinks=true,citecolor=blue]{hyperref}

@@

wwr_mle_raw <- c(params_fixed, 
             "beta1" = 11.48,
             "beta2" = 0.25,
             "beta3" = 0.35,
             "phi" = 0.14,
             "beta11" = 0.16,
             "sigma_q" = 0.021,
             "sigma_xi" = 66.89)

@@

bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_mif_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
                    beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
                    S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
                    S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
                    S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
  mifs_global_01
})

bake(file = paste0(out_dir, "mif2_maximization/round_01/ovov_el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/ovov_el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"mif2_maximization/round_02/ovov_mif_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                    beta11=0.01*sd_02,phi=0.01*sd_02,
                    sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                    S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                    S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                    S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/ovov_el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/round_03/ovov_mif_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                    beta11=0.01*sd_03,phi=0.01*sd_03,
                    sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                    S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                    S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                    S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
mif_el <- readRDS(paste0(out_dir,"mif2_maximization/round_03/ovov_el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- mif_el[which.max(mif_el$logLik),]
@

@@

bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
prop <- 420

@@

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_100.rds"),{
  pal_log_lik <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    realdat <- ovov_sim_100[[i]] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                     c(initial_guess[1],initial_guess[2],initial_guess[3],
                                       initial_guess[4],initial_guess[5]), 
                                     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                     norm_par = c(0.07,initial_guess[6]), 
                                     t(prop), NP_MIF,1)
    pal_log_lik[i] <- lik_list$log_lik
  }
  pal_log_lik
})

@@

<<pfilter_org, echo=FALSE, message=FALSE, warning=FALSE>>=
dat[-1, ] |>
  pomp(t0=0,
       time="time",
       rprocess=discrete_time(rproc,delta.t=1/4), 
       rinit=rinit_wwr,
       dmeasure=dmeas,
       rmeasure=rmeas,
       accumvars=c("H1","H2","H3"),
       statenames=c("S1","S2","S3",
                    "I1","I2","I3",
                    "R1","R2","R3",
                    "H1","H2","H3",
                    "q1","q2","q3"),
       paramnames = names(wwr_mle_raw)
  ) -> m1

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_org.rds"),{
  registerDoFuture()
  plan(multicore, workers = ncores)
  foreach(i=1:ncores, .combine=c,
          .options.future=list(seed=998468235L)
  ) %dofuture% {
    pfilter(m1,Np=NP_MIF,params=wwr_mle_raw)
  } -> pfs
  pfs
})

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_list_cond_lik.rds"),{
  list_cond_lik <- list()
  registerDoFuture()
  plan(multicore, workers = ncores)
  for(k in 1:ncores){
    list_cond_lik[[k]] <- pfs[[k]]@cond.logLik
  }
  list_cond_lik
})
@

<<palfilter_org, echo=FALSE>>=
load("real_rotavirus_metadata.Rdata")
prop <- 420

### palfiter
registerDoFuture()
plan(multicore, workers=ncores)

Sys.time()
bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_org.rds"),{
  pal_lik_list_org <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
                                   c(initial_guess[1],initial_guess[2],
                                     initial_guess[3],initial_guess[4],initial_guess[5]), 
                                   gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
                                   norm_par = c(0.07,initial_guess[6]), 
                                   t(prop), NP_MIF,1)
  pal_lik_list_org
})
Sys.time()
@

<<plot1, echo=FALSE, fig.height=4, fig.width=5.5>>=
# Create your data
time <- 1:416
ncores <- length(list_cond_lik)

# Combine the data into a single dataframe for SMC
data_list <- lapply(1:ncores, function(i) {
  data.frame(time = time, cond_loglik = list_cond_lik[[i]], method = "SMC")
})

# Combine all the SMC data frames into one
df_smc <- bind_rows(data_list)

# Create the PAL data frame
df_pal <- data.frame(time = time, cond_loglik = pal_lik_list_org$ll_storage, method = "PAL")

# Combine both data frames
df <- bind_rows(df_smc, df_pal)

# Plot with ggplot
ggplot(df, aes(x = time, y = cond_loglik, group = interaction(method, factor(1:ncores)), color = method)) +
  geom_line(data = filter(df, method == "SMC"), color = "blue") +
  geom_line(data = filter(df, method == "PAL"), color = "red") +
  theme_minimal() +
  labs(title = "Conditional Log Likelihood Over Time",
       x = "Time",
       y = "Conditional Log Likelihood",
       color = "Method") +
  theme(legend.position = "bottomright")

@

\includegraphics[width=4.5in]{plot1}

@@

\begin{table}[ht] % The placement specifier can be [h!tbp]
\centering % Centers the table
\caption{log-likelihood at the MLE of OvOv model for the real rotavirus data, computed using two filtering methods, employed 36 replicates and 50,000 particles} % This is effectively the table title
\label{tab:ovovrealdata} % For referencing this table elsewhere in your text
\begin{tabular}{|c c c|} 
 \hline
  & PAL & SMC  \\
  \hline
 OvOv & \Sexpr{pal_lik_list_org$log_lik} & $\Sexpr{logmeanexp(sapply(pfs, logLik))}$ \\ 
 \hline
\end{tabular}
\end{table}

@@
